---
title: "Stat 440 0111"
author: "Sangjun Pyo"
date: "5/19/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Inverse Transformation Sampling

Simulating random variables is the foundation of countless methods in statistical computing. 
Many things we do in this class will rely on the ability to simulate random variables. 
By the end of this lesson, you should be able to:

• Understand how computers generate approximately random numbers.
• Learn how to use inverse transformation to generate samples from continuous and discrete random       variables.
• Apply these techniques to R examples.

## Uniform Distribution

```{r }
# compute CDF of Unif(0, 1)
punif(c(-2, 0, 0.27, 0.64, 1, 2), min=0, max=1)

# compute PDF of Unif(0, 1)
dunif(c(-2, 0, 0.27, 0.64, 1, 2), min=0, max=1)

# compute quantile of Unif(0, 1)
qunif(c(-2, 0, 0.27, 0.64, 1, 2), min=0, max=1)

# Uncomment the line below and run once
# if you don't have ggplot2 installed.

# install.packages("ggplot2")

library(ggplot2)
# plot CDF of Unif(0, 1) distribution
xset <- seq(-.5, 1.5, .001)
udat <- data.frame(x=xset,y=punif(xset, min=0, max=1))
ggplot(data=udat,mapping=aes(x=x, y=y)) +
  geom_line() + ylab("CDF") + ggtitle("CDF of Unif(0, 1)")
```

## Pseudo-Random Numbers

Simulating random numbers on computers can be hard for two key reasons:

1. Computers are deterministic: computer code follows written instructions exactly, 
and it’s impossible for computers to generate truly random numbers.

2. Computers are finite: computers can only approximate real numbers, and
they cannot generate exact real numbers (e.g., irrational numbers like $\pi$)

To overcome these issues, computers use pseudo-random number generation (PRNG). 
These are deterministic algorithms that start with an initial value (called a seed value) and deterministically output numbers that are approximately random. 
Most PRNG algorithms generate uniform random numbers;
here is R’s PRNG to generate from Unif(0, 1).

```{r }
# set PRNG seed
set.seed(123)
# generate 3 pseudo-random numbers from Unif(0, 1)
runif(3, min=0, max=1)

```

The PRNG output is entirely determined by the seed value. If we use the
same seed, we get the same output:

```{r}
set.seed(123)
runif(3, min=0, max=1)
```

If we use a different seed, we get a different output:

```{r}
set.seed(456)
runif(3, min=0, max=1)
```


## Inverse Transformation Method

For the purposes of this class, we will start our analyses assuming the ability
to generate pseudo-random numbers from Unif(0, 1). This allows us to use
the inverse transformation method, the focus of this lesson.

## Example 1: Exponential Distribution

```{r}
# Step 1: generate uniform samples
unif_samples <- runif(1000, min=0, max=1)
# Step 2: apply inverse transformation
exp_samples <- -log(1 - unif_samples) / 2

# plot histogram of samples along with true density
ggplot(data=data.frame(x=exp_samples)) +
  # create histogram of samples
  geom_histogram(mapping=aes(x=x, y=..density..),
                fill="white", color="black", bins=30) +
  # add true density using 'dexp' function
  geom_line(data=data.frame(x=seq(0, 3, .001),
                            y=dexp(seq(0, 3, .001), rate=2)),
            mapping=aes(x=x, y=y),color="blue") +
  ggtitle("Exp(lambda=2) simulation vs truth")
```

## Example 2: Vaccination Rate

```{r}
# probability mass function (PMF) as a vector
dosage_probs = c(.218, .143, .437, .202)

# calculate cumulative sum of PMF values
cumsum(dosage_probs)

```

We plot the true dosage CDF below. Note that our CDF is NOT continuous,
so it is not directly invertible.

```{r}
# plot the true CDF
ggplot(data=data.frame(x=c(-1, 0, 1, 2, 3, 4),
                        y=cumsum(c(0, dosage_probs, 0))),
      mapping=aes(x=x,y=y)) +
  geom_step(direction="hv") +
  ggtitle("Dosage count CDF") +
  ylab("CDF")

```

However, we can still use inverse transformation. First, we define our CDF
and quantile functions as piece-wise linear functions:

```{r}
dosage_cdf = function(x) {
  #' CDF for dosage count distribution
  #' @param x input value
  if (x < 0) {
    0
  } else if (x < 1) {
    .218
  } else if (x < 2) {
    .361
  } else if (x < 3) {
    .798
  } else {
    1
  }
}

dosage_inv_cdf = function(v) {
  #' Inverse CDF for dosage count distribution
  #' @param v input value
  if (v < .218) {
    0
  } else if (v < .361) {
    1
  } else if (v < .798) {
    2
  } else {
    3
  }
}
```

In the examples above, we wrote out all the cases explicitly. Here’s another
implementation using which(); this may be helpful when there are too many
cases to enumerate explicitly

```{r}
dosage_inv_cdf_alt = function(v) {
  #' Inverse CDF for dosage count distribution without cases
  #' @param v input value
  which(v < cumsum(dosage_probs))[1] - 1
}
```

Next, we calculate the inverse CDF by applying our function to uniform
samples:

```{r}
# step 1: generate uniform samples
dosage_unif_sams = runif(1000, min=0, max=1)

# step 2: apply inverse transformation
dosage_dist_sams = sapply(dosage_unif_sams,dosage_inv_cdf)

# summarize results in a table
table(dosage_dist_sams)/1000
```

This distribution based on simulations is very similar to our target distribution, so the method works as expected.